Bees need larger brains to thrive in urban environments

### 2023-06-12 Manuscript/Ms.Rmd
text

---
title: "Bees need larger brains to thrive in urban environments"
output:
  pdf_document:
    fig_caption: yes

link-citations: yes
linkcolor: RoyalBlue
urlcolor: RoyalBlue
fontsize: 12pt

bibliography: 
  - references.bib

header-includes:
- \usepackage{orcidlink}
- \usepackage{caption,setspace}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{float} 
- \floatplacement{figure}{H}

csl: BiologyLetters.csl

---

@@

d = raw %>% 
group_by(Species) %>% 
summarise(individuals = n())

@@

The rapid conversion of natural habitats to anthropogenic landscapes is threatening insect pollinators worldwide, raising concern on the consequences for their fundamental role as plant pollinators in natural and agricultural systems. However, not all pollinators are negatively affected by habitat conversion, as some find in anthropogenic landscapes appropriate resources to survive and reproduce. Why some animals can thrive in anthropogenic environments that most species cannot tolerate remains insufficiently understood, but it might in part reflect enhanced behavioural plasticity to exploit new resource opportunities. This theory -known as the cognitive buffer hypothesis- has received ample support in studies of vertebrates. However, it is less obvious whether the tiny brain of insects can similarly buffer individuals against the many challenges of anthropogenic habitats. To address this issue, we measured brains for `r sum(d$individuals)` individuals from `r nlevels(factor(unique(d$Species)))` species of bees, the most important group of pollinators, and evaluated the association of their brain size with detailed information on habitat preference. Our analyses revealed that bees that prefer forested or agricultural habitats had smaller brains relative to their body size, but those who prefer urban habitats tended to have larger relative brain size. In addition, urban bees also tend to have larger body sizes, and hence, show larger absolute brain sizes. Our results provide the first empirical support of the cognitive buffer hypothesis in invertebrates, suggesting that a large brain in bees could confer behavioural advantages to tolerate urban environments.

Keywords: relative brain size, habitat preference, Apoidea, urbanization.

@@

Pollinators deliver a fundamental ecosystem service on which the earth's vegetation and human economy depend [@ollerton2017]. Regrettably, there is increasing evidence of recent declines in pollinator populations [@bartomeus2019; @potts2010; @scheper2014]. One of the main contributing factors to the current pollinator declines is the alteration and loss of their habitat due to human activity [@harrison2015; @harrison2018]. Anthropogenic landscapes present new challenges for survival and reproduction of organisms, increasing their risk of extinction [@winfree2011; @brown2009 ; @harrison2019]. Yet, not all pollinator species are negatively affected by land use change. For instance, some bee species are able to tolerate human-altered environments  [@biesmeijer2006; @bartomeus2013; @collado2019] or even to thrive such as *Bombus terrestris* [@rasmont2008]. However, we still lack generalizable information about the functional and ecological consequences of inhabiting these novel environments [@buchholz2020].

@@

Insects have been long associated with the most basic types of learning due to their miniature brain [@giurfa2015; @perry2017]. However, numerous studies have shown that bees and other insects have sophisticated cognition [@chittka2009; @avargues2011; @sheehan2011]. For instance, some social insects can learn by observing other individuals how to use tools to solve complex tasks [@alem2016]. In addition, there is also evidence for substantial variation across species in brain size—both in absolute terms and relative to body size— with species that have larger brains also exhibiting enhanced cognitive performance [@collado2021]. Therefore, we can ask whether the varying success of insects in human-altered habitats might be explained by variation in brain size.

Here, we report the first test of the “cognitive-buffer hypothesis“ in insects. Our test is based on a unique database of brain measures for `r nlevels(factor(unique(d$Species)))` European and North American bee species. By means of detailed information of geographic distribution (GBIF; http://www.gbif.org/), we characterise habitat preferences for all the species, and use a phylogenetically-informed comparative analysis to assess whether bees that proliferate in human-altered habitats have enlarged brains compared to those that avoid them.

@@

Our dataset contains measurements of brain and body size for bee specimens captured on flowers by hand netting in different areas of the East Coast of the United States and Europe (Spain and the Netherlands). The dataset includes information of `r sum(d$individuals)` female individuals from `r nlevels(factor(unique(d$Species)))` species that represent `r nlevels(factor(unique(tax$family)))` families and `r nlevels(factor(unique(tax$genus)))` genera. We only considered female specimens because: (i) females are involved in a greater diversity of tasks than males, and hence are expected to experience greater selective pressures from the environment; and (ii), there are evident behavioural and morphological differences between sexes that are associated with the size and structure of the brain [@streinzer2013; @roselino2015]. Brain size was measured as the weight of fixed brains and body size as intertegular span (see supplementary text). Because brain size scales allometrically with body size, we considered for our analysis the size of the brain relative to the body.

We estimated brain size relative to body size through modelling the log-transformed association between brain weight and intertegular span. For this, we specified a Bayesian linear model with the function *brm* from the package *brms* version 2.18 [@burkner2017] where brain weight was the response variable and intertegular span the predictor. To correct for the non-independent evolutionary histories among species, we included the phylogenetic covariance matrix as a random factor. Lastly, we extracted the differences between the observed and predicted values (i.e., residuals) that represent the relative brain size. High values indicate larger brains than expected for their body size and low values denote smaller brains than expected. We found that brain size scaled allometrically with body size (Bayesian R$^2$ = `r r2`) and that this relationship was constrained by the evolutionary history of the species (phylogenetic signal of relative brain size, $\lambda$ = `r lambda`; *P* < 0.01). However, there were considerable variability in relative brain size within and across the different taxonomic groups (i.e., genera and subfamily level; Figure S1; Figure 1A)

@@

We assigned a habitat type to each GBIF occurrence by merging land cover information with the georeferenced records of species occurrences. The land cover classification was obtained from the 2006 online inventories of the National Land Cover Database (NLCD) for United States and the Corine Land Cover (CLC) for Europe. After downloading these inventories as raster files, we used the functions *rast* and *extract* from the *Terra* package version 1.6-41 [@hijmans2022] to read and obtain the cover classification of the different georeferenced records, respectively. To simplify the interpretation and conduct a joint analysis for both occurrence datasets, we divided the resulting cover classes into three single categories: (i) natural, (ii) agricultural and (iii) urban (see Tables S1 and S2 for details).

## (d) Construction of the phylogenetic tree

We constructed a phylogeny with all studied species by using a previously published genus-level bee phylogeny [@hedtke2013]. For this, we first read the backbone phylogeny with the function *read.tree* and drop the missing genera with function *drop.tip* from the package *ape* version 5.6-2 [@paradis2019]. Then, we selected as an outgroup a wasp genera (*Tachysphex*) and recalculated distances to the root with the function *chronos* from *ape*. Species tips were added to genera nodes as polytomies of equal branch length relative to the genera branch length using the phytools package version 1.2-0 [@revell2012]. From the resulting tree, we created a scaled phylogenetic covariance matrix with the function *inverseA* from the package *MCMCglmm* version 2.33 [@hadfield2010].

## (e) Analysis

Habitat preference was estimated by comparing species occurrences per habitat with the ones that we would expect if they occur at random. For this, we generated 10,000 randomised matrices based on the occurrence matrix (i.e., species in rows and habitats in columns) with the function *nullmodel* from the package *bipartite* version 2.16 [@dormann2008]. We implemented this function with the method 'r2dtable', which maintains row and column sums constant by using Patefield's algorithm [@patefield1981]. This maintains the proportional dominance of the species and habitats constant but reshuffles their associations. We then estimated the percentage of simulated occurrences per species and habitat that were under the observed ones (i.e., percentile). Lastly, we considered that a bee species exhibited a “high preference” for any of the three studied habitats when the number of  occurrences observed in this habitat exceeded  the 80th percentile of the values obtained from the simulations. On the contrary, the species was considered to exhibit a “low preference” for the habitat when the observed occurrences were below the 20th percentile. To better understand if our findings are affected by the evolutionary histories of the species, we estimated the phylogenetic signal of relative brain size and habitat preference across habitats for our set of species with the help of the function *phylosig* from the *phytools* package.

To evaluate how the association between habitat preference and relative brain size changed by habitat type, we modelled with Bayesian generalised linear models their association. For this, we first joined the resulting habitat preference and relative brain size datasets by species and extracted those with statistical habitat preference on any of these three habitats. Because we assessed habitat preference as binary (low or high), we specified a Bernouilli distribution where habitat preference was the response variable and relative brain size the predictor. Again, we included the phylogenetic covariance matrix as a random factor. Moreover, we also explored how habitat preference changed by brain weight and intertegular span independently. For this, we conducted two analogous models with these two different predictors. In addition, we also investigated the different trends for the United States and Europe separately.

All our models were run with 4,000 iterations with previous 1,000 warm up iterations, usingnon-informative or weakly informative priors [@burkner2017]. Further, the different posterior predictive checks were conducted with the function *pp_check*, also from the *brms* package. All our analyses were undertaken in R version 4.0.5 [@r2021], and all data processing and graphics were done with the set of packages from the tidyverse version 1.3.0 [@tidyverse19].

@@

Habitat preference varied substantially across species (Figure 1B) but showed none (i.e., preference for agricultural and rural habitats) or moderate phylogenetic signal (i.e., preference for natural habitat; $\lambda$ = `r lambda_nat`; *P* = `r p_nat`). In general, most bee species preferred one or two habitat types but more rarely occurred indistinctly in the three habitats (Figure 1B). The most preferred habitats were the agricultural and natural ones with `r pref_agr` and `r pref_nat` species over 80% of the values from null models, respectively (Figure 1B and Figure S3). In contrast, most species avoided or had low preference for urban habitats (`r pref_urb1` species under the 20th percentile) and just `r pref_urb2` species showed high preference for this habitat type (Figure 1B and Figure S3). The resulting distribution of habitat preferences for each of the habitats analysed followed a zero-one inflated beta distribution (Figure S3) but note that for our analyses we only modelled the extremes of the distribution (i.e., species with statistical avoidance or preference). Hence, there were high frequencies of habitat preferences close to 0 or 1 but low frequencies of intermediate values between 0 and 1.

@@

Although our model evaluating how relative brain size was associated with habitat preference explained little of the overall variance (Bayesian R$^2$ = `r r2_main`), we found evident differences when we explored this association by habitat type (Figure 2A). Specifically, we found that bees with larger brains relative to body size showed a higher preference for urban habitats than bees with smaller relative brain sizes (Figure 2A). Contrarily, bees with smaller relative brains showed higher preference in natural and agricultural habitats than bees with larger relative brain sizes (Figure 2A). Although the models of the association between brain size and intertegular span captured also little of the overall variance (Bayesian R$^2$ = `r r2_brain`; Bayesian R$^2$ = `r r2_its`, respectively), we found more marked differences in this case between habitat types (Figures 2B and 2C). Specifically, we found that bees with larger brains and body sizes tend to appear more in urban habitats and bees with smaller brains and body sizes are more common in natural and agricultural ones (Figures 2B and 2C). These findings were consistent with the analogous analyses separated by geographical regions (United States and Europe; Figure S4).

@@

Bees with a preference for inhabiting urban habitats tended to have larger brains relative to their body, and those with forest or agricultural preferences had relatively smaller brains. These results are in line with the cognitive buffer hypothesis [@sol2009], which predicts that novel environments should select for larger brains. In urban environments, individuals are frequently exposed to a variety of new challenges such as novel resources or human disturbances that can change fast in time and space [@wheater2002]. In such scenarios, a large brain could confer the behavioural flexibility needed to efficiently exploit novel resources or avoid novel predators. Our findings are also consistent with previous evidence found in vertebrates that having a large brain size can allow survival in urban environments [@santini2019; @sayol2020b], and suggest that, in bees, urban dwellers may need to have better cognitive abilities and flexibility in their behaviour than forest or agricultural bees.

The exact mechanism that allows large-brained bee species to prefer urban habitats is still unclear. However,  there are some examples of novel behaviours that might be needed to survive in urban environments, such as the use of human-made materials to nest [@allasino2019] or being able to collect pollen and nectar from exotic flowers [@lowenstein2019]. Despite the lack of broad consensus of the functional consequences of urbanisation [@buchholz2020], there is support to the idea that cities select against small body sizes [@ahrne2009; @theodorou2021]. We found support for this functional consequence with occurrence data from two different continents and over 100 bee species. This may be a consequence of the patchy resource distribution as the foraging distance in bees is positively associated with body size [@greenleaf2007; @kendall2019]. Further, the patchiness of urban resources might also favour species that are already adapted to look for target resources (i.e. specialist bees), which tend to have relatively larger brains [@sayol2020]. However, this hypothesis would need to be properly tested for bees, as there are shreds of evidence that associate increasing urbanisation with wider ecological niches [@geslin2013; @buchholz2020; @prendergast2022].

Our analyses focused on brain size mainly for reasons of data availability. The use of brain size is based on the existence of previous evidence that enlarged brains enhance cognitive performance in bees [@collado2021] and is justified because the brain is less subject
to measurement error or context-dependent biases than are experimental measures of cognition [@kamil1987]. However, the use of brain size as a proxy for cognitive performance is not exempt from criticism [@chittka2009; @lihoreau2012]. Thus, finer brain measurements like neuropils size or mushroom bodies [the suspected centres of cognitive processes in bees\; @barth1997; @withers1995; @fahrbach1995] are likely to help improve our understanding of bee neurobiology. Overall, our findings highlight the importance of the behavioural responses for understanding the dynamics of insect populations in altered environments, and stresses the need to avoid viewing them as passive agents of external factors.

@@

@@

Brain size was measured as the weight of the fixed brain for each specimen. First, all collected individuals were kept in cold and sent to the laboratory where they were identified by expert taxonomists and anesthetized in cold. Then, heads were removed with a scalpel and placed in a fixative solution (4% paraformaldehyde in 0.1 M phosphate-buffered saline -PBS- pH = 7.4) to avoid degradation. Subsequently, brains were extracted from the head capsule and placed on a petri dish after removing the retina from the optic lobes and cleaning the entire brain of all tracheae and fat bodies. By doing this, we ensured that our brain measurements mainly accounted for neural tissue. We then placed the brains on a small piece of Parafilm® and dried the exceeding fixative solution with Kimwipes® tissues. Then, within 4 seconds of the liquid removal, we weighted the brains with the help of a microbalance Sartorius Cubis®. Finally, we calculated the average weight for the species with more than one specimen. Because weighing bee brains is a highly error-prone task, we excluded values that were under or above 1.5 times the interquartile range for each species.

>>>>

---
title: "Bees need larger brains to thrive in urban environments"
output:
  word_document: default
  pdf_document:
    fig_caption: yes
link-citations: yes
linkcolor: RoyalBlue
urlcolor: RoyalBlue
fontsize: 12pt
bibliography: references.bib
always_allow_html: true
header-includes:
- \usepackage{orcidlink}
- \usepackage{caption,setspace}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{float}
- \floatplacement{figure}{H}
csl: BiologyLetters.csl
---

@@

d = raw %>% 
group_by(Species) %>% 
summarise(individuals = n())
#Load preferences data
preferences_usa = readr::read_csv("../Data/USA_data/preferences_usa.csv") 
preferences_eu = readr::read_csv("../Data/Europe_data/preferences_europe.csv") 
#Bind rows
pref = bind_rows(preferences_usa, preferences_eu)
s = unique(pref$Species)

raw = raw %>% filter(Species %in% s)

@@

The rapid conversion of natural habitats to anthropogenic landscapes is threatening insect pollinators worldwide, raising concern on the consequences for their fundamental role as plant pollinators. However, not all pollinators are negatively affected by habitat conversion, as certain species find in anthropogenic landscapes appropriate resources to persist and proliferate. The reason why some species thrive in anthropogenic environments while most find them inhospitable remains poorly understood. The cognitive buffer hypothesis, widely supported in vertebrates but untested in insects, offers a potential explanation. This theory  suggests that species with larger brains have enhanced behavioural plasticity, enabling them to confront and adapt to novel challenges. To investigate this hypothesis in insects, we measured brains for `r nrow(raw)` individuals from `r nlevels(factor(unique(raw$Species)))` species of bees, and evaluated the association of their brain sizes with detailed information on habitat preferences. Our analyses revealed that bees that prefer urban habitats had larger brains relative to their body size than those who prefer forested or agricultural habitats. Additionally, urban bees exhibited larger body sizes and, consequently, larger absolute brain sizes. Our results provide the first empirical support to this hypothesis in invertebrates, suggesting that a large brain in bees could confer behavioural advantages to tolerate urban environments.

Keywords: relative brain size, habitat preference, Apoidea, urbanisation, pollinators

@@

Pollinators deliver a fundamental ecosystem service on which the earth's vegetation and human economy depend [@ollerton2017]. Regrettably, there is increasing evidence of recent declines in pollinator populations [@bartomeus2019; @potts2010; @scheper2014]. One of the main contributing factors to the current pollinator declines is the alteration and loss of their habitat due to human activity [@harrison2015; @harrison2018]. Anthropogenic landscapes present new challenges for survival and reproduction of organisms, increasing their risk of extinction by maldaptation [@winfree2011; @brown2009 ; @harrison2019]. Yet, not all pollinator species are negatively affected by land use change. For instance, some bee species are able to tolerate human-altered environments [@biesmeijer2006; @bartomeus2013; @collado2019] or even to thrive [e.g., _Bombus terrestris_\; @rasmont2008; @theodorou2020].

@@

Insects have been long associated with the most basic types of learning due to their miniature brain [@giurfa2015; @perry2017]. However, numerous recent studies have shown that bees and other insects have sophisticated cognition that goes beyond simple associative learning or conditioning [@chittka2009; @avargues2011; @sheehan2011]. Some of these more complex forms of cognition involve the use of simple tools, attention, social learning or metacognitive processes [@giurfa2015; @perry2017]. In addition, there is also evidence for substantial variation across species in brain size—both in absolute terms and relative to body size— with species that have larger brains also exhibiting enhanced cognitive performance [@collado2021]. Therefore, we can ask whether the varying success of insects in human-altered habitats might be explained by variation in brain size.

Here, we report the first test of the “cognitive-buffer hypothesis“ in insects. Our test is based on a unique database of brain measures for `r nlevels(factor(unique(d$Species)))` European and North American bee species. By means of detailed georeferenced information of species occurrences (GBIF; http://www.gbif.org/), we characterise habitat preferences for all the species, and use a phylogenetically-informed comparative analysis to assess whether bees that proliferate in human-altered habitats have enlarged brains compared to those that avoid them.

@@

Our dataset contains measurements of brain and body size for bee specimens captured on flowers by hand netting in different areas of the East Coast of the United States and Europe (Spain and the Netherlands). The dataset includes information of `r nrow(raw)` female individuals from `r nlevels(factor(unique(raw$Species)))` species that represent `r nlevels(factor(unique(tax$family)))` families and `r nlevels(factor(unique(tax$genus)))` genera. We only considered female specimens because: (i) they are involved in a greater diversity of tasks than males, and hence are expected to experience greater environmental pressures; and (ii), have brains that are structurally and functionally different from those of males [@streinzer2013; @roselino2015]. Brain size was measured as the weight of fixed brains [@sayol2020] and body size as intertegular span [@kendall2019\; see supplementary text]. Given that brain size scales allometrically with body size [@sayol2020], we considered for our analysis the size of the brain relative to the body.

Following @sayol2020, we estimated relative brain size as the residuals of a log-log phylogenetic linear model of brain mass against body size, using the Bayesian approximation implemented in the function *brm* from the package *brms* version 2.18 [@burkner2017]. The phylogeny was built with the help of a previously published genus-level phylogeny and it was processed with the help of the packages ape 5.6-2 [@paradis2019], phytools 1.2-0 [@revell2012] and MCMCglmm version 2.33 [@hadfield2010]. High values of relative brain size indicate larger brains than expected for their body size and low values denote smaller brains than expected. Our bee dataset showed a strong allometric relationship between brain and body size (Bayesian R$^2$ = `r r2`) that was constrained by the evolutionary history of the species (phylogenetic signal of relative brain size, $\lambda$ = `r lambda`; *P* < 0.01). However, we found considerable variability in relative brain size within and across the different taxonomic groups (i.e., genera and subfamily level; Figure S1; Figure 1A).

@@

We assigned a habitat type to each GBIF occurrence by merging land cover information with the georeferenced records of species occurrences. The land cover classification was obtained from the 2006 online inventories of the National Land Cover Database (NLCD) for United States and the Corine Land Cover (CLC) for Europe. After downloading these inventories as raster files, we used the functions *rast* and *extract* from the *Terra* package version 1.6-41 [@hijmans2022] to read and obtain the cover classification of the different georeferenced records, respectively. To simplify the interpretation and conduct a joint analysis for both regions, we divided the resulting cover classes into three single categories: (i) natural, (ii) agricultural and (iii) urban (see Tables S1 and S2 for details).

## (d) Analysis

Habitat preferences were estimated by assessing whether the occurrences of species in a given habitat were more frequent than expected by chance. For this, we generated 10,000 randomised matrices based on the occurrence matrix (i.e., species in rows and habitats in columns) with the function *nullmodel* from the package *bipartite* version 2.16 [@dormann2008]. We used the method 'r2dtable', which maintains row and column sums constant by using Patefield's algorithm [@patefield1981]. This maintains the proportional dominance of the species and habitats constant but reshuffles their associations. We then estimated the percentage of simulated occurrences per species and habitat that were under the observed ones (i.e., percentile). Lastly, we considered that a bee species exhibited a “high preference” for any of the three studied habitats when the number of  occurrences observed in this habitat exceeded  the 80th percentile of the values obtained from the simulations. On the contrary, the species was considered to exhibit “low preference or avoidance” for the habitat when the observed occurrences were below the 20th percentile. To better understand if our findings are affected by the evolutionary histories of the species, we estimated the phylogenetic signal of relative brain size and habitat preference across habitats for our set of species with the help of the function *phylosig* from the *phytools* package.

To evaluate how the association between habitat preference and relative brain size changed by habitat type, we modelled with Bayesian generalised linear models their association. For this, we first joined the resulting habitat preference and relative brain size datasets by species and extracted those with statistical habitat preference on any of these three habitats. The resulting distribution of habitat preferences for each of the habitats analysed followed a zero-one inflated beta distribution (Figure S3), indicating that there were high frequencies of habitat preferences close to 0 or 1 but low frequencies of intermediate values between 0 and 1. Hence, in our analyses we only modelled the extremes of the distribution (i.e., species classified as avoiding or preferring a given habitat). Because we assessed habitat preference as binary (low or high), we specified a Bernouilli distribution where habitat preference was the response variable and relative brain size the predictor. Again, we included the phylogenetic covariance matrix as a random factor. Moreover, we also explored how habitat preference changed by brain weight and intertegular span independently. For this, we conducted two analogous models with these two different predictors. In addition, we also investigated the different trends for the United States and Europe separately.

All our models were run with 4,000 iterations with previous 1,000 warm up iterations, using non-informative or weakly informative priors [@burkner2017]. Further, the different posterior predictive checks were conducted with the function *pp_check*, also from the *brms* package. All our analyses were undertaken in R version 4.0.5 [@r2021], and all data processing and graphics were done with the set of packages from the tidyverse version 1.3.0 [@tidyverse19].

@@

Habitat preference varied substantially across species (Figure 1B) but showed none (i.e., preference for agricultural and rural habitats) or moderate phylogenetic signal (i.e., preference for natural habitat; $\lambda$ = `r lambda_nat`; *P* = `r p_nat`). In general, most bee species preferred one or two habitat types but more rarely occurred indistinctly in the three habitats (Figure 1B). The most preferred habitats were the agricultural and natural ones with `r pref_agr` and `r pref_nat` species over 80% of the values from null models, respectively (Figure 1B and Figure S3). In contrast, most species avoided urban habitats (`r pref_urb1` species under the 20th percentile) and just `r pref_urb2` species showed high preference for this habitat type (Figure 1B and Figure S3).

@@

Although the model evaluating how relative brain size was associated with habitat preference explained little of the overall variance (Bayesian R$^2$ = `r r2_main`), we found evident trends when we explored this association by habitat type (Figure 2A). Specifically, we found that bees with larger brains relative to body size showed a higher preference for urban habitats than bees with smaller relative brain sizes (Figure 2A). Contrarily, bees with smaller relative brains showed higher preference in natural and agricultural habitats than bees with larger relative brain sizes (Figure 2A). The models of the association between absolute brain size and intertegular span with habitat preference captured also little of the overall variance (Bayesian R$^2$ = `r r2_brain`; Bayesian R$^2$ = `r r2_its`, respectively). However, we found again marked differences between habitat types (Figures 2B and 2C). Specifically, we found that bees with larger brains and body sizes tend to appear more in urban habitats and bees with smaller brains and body sizes are more common in natural and agricultural ones (Figures 2B and 2C). These findings were consistent with the analogous analyses separated by geographical regions (United States and Europe; Figure S4).

@@

Using data from two different continents and 89 bee species, we find that bees with a preference for inhabiting urban habitats tended to have larger brains relative to their body while those with forest or agricultural preferences had relatively smaller brains. These results are in line with the cognitive buffer hypothesis [@sol2009], which predicts that a large relative brain should provide enhanced behavioural plasticity to persist and thrive in novel environments. In urban environments, individuals are frequently exposed to a variety of new challenges such as novel resources or human disturbances that can change fast in time and space [@wheater2002]. In such scenarios, a large brain could confer the cognitive flexibility needed to efficiently exploit novel resources while avoiding risks. While the exact mechanisms are unclear, we expect cognitive flexibility to be essential in a variety of contexts such as the use of human-made materials for the nest [@allasino2019] or the need to collect pollen and nectar from exotic flowers [@lowenstein2019].

Also in line with the cognitive buffer hypothesis [@ducatez2020], we find that bees that proliferate in urban habitats have bigger bodies and larger absolute brains. Larger species obtain greater net benefits from exploring and learning, especially in heterogeneous environments [@bateman2012; @theodorou2021]. The patchiness of urban resources seems indeed to favour larger body sizes, as suggests the observation that foraging distance is positively associated with body size in bees [@greenleaf2007; @kendall2019]. There are also shreds of evidence that associate increased tolerance to urbanisation with wider ecological niches [@geslin2013; @buchholz2020; @prendergast2022], and both larger bodies and relative brains are thought to be features that can facilitate broader diets and niche expansion. Further understanding all these processes represent an important avenue of future research.

Overall, our findings support and expand previous evidence found in vertebrates that having a large brain size can allow survival in novel environments [@santini2019; @sayol2020b], highlighting that a cognitive buffer is possible even with tiny brains. The use of the entire brain size as a proxy for cognitive performance is not exempt from criticism [@chittka2009; @lihoreau2012]. Our analyses focused on brain size mainly for reasons of data availability. The use of brain size is justified by the existence of previous evidence that enlarged brains enhance cognitive performance in bees [@collado2021]. Moreover, brain size is less subject to measurement error or context-dependent biases in comparison with other experimental measures of cognition [@chittka2009; @lihoreau2012]. Although the use in the future of finer measures like neuropils size or mushroom bodies [the suspected centres of cognitive processes in bees\; @barth1997; @withers1995; @fahrbach1995] is likely to help improve our understanding of bee cognitive abilities. Our findings highlight the importance of behavioural responses for understanding the dynamics of insect populations in altered environments and stresses the need to avoid viewing them as passive agents of external pressures.

@@

Ministerio de Economía y Competitividad, Gobierno de España, Grant/Award Number: CGL2013-47448-P

@@

Brain size was measured as the weight of the fixed brain for each specimen. First, all collected individuals were kept in cold and sent to the laboratory where they were identified by expert taxonomists and anaesthetised in cold. Then, heads were removed with a scalpel and placed in a fixative solution (4% paraformaldehyde in 0.1 M phosphate-buffered saline -PBS- pH = 7.4) to avoid degradation. Subsequently, brains were extracted from the head capsule and placed on a petri dish after removing the retina from the optic lobes and cleaning the entire brain of all tracheae and fat bodies. By doing this, we ensured that our brain measurements mainly accounted for neural tissue. We then placed the brains on a small piece of Parafilm® and dried the exceeding fixative solution with Kimwipes® tissues. Then, within 4 seconds of the liquid removal, we weighted the brains with the help of a microbalance Sartorius Cubis®. Finally, we calculated the average weight for the species with more than one specimen. Because weighing bee brains is a highly error-prone task, we excluded values that were under or above 1.5 times the interquartile range for each species.
